\Title

### 2013-01-28 synmut.tex
histograms for SHAPE instead of cumulative
combed through the text. Discussion and captions still need work

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ARTICLE ABOUT FATE OF SYNONYMOUS MUTATIONS IN HIV
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[rmp, twocolumn]{revtex4}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage{amsmath,amsfonts,amssymb,eucal,eurosym,textcomp}
\usepackage{color}
\usepackage{graphicx}
\usepackage[caption=false]{subfig}
\usepackage{natbib}
\usepackage{pslatex}
\usepackage[colorlinks,linkcolor=red,citecolor=red]{hyperref}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\graphicspath{{./figures/}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\comment}[1]{\textit{\textcolor{red}{#1}}}
\newcommand{\mut}{\mu}
\newcommand{\mfit}{\langle F\rangle}
\newcommand{\mexpfit}{\langle e^{F}\rangle}
\newcommand{\ox}{r}
\newcommand{\co}{\rho}
\newcommand{\gt}{g}
\newcommand{\locus}{s}
\newcommand{\locuspm}{t}
\newcommand{\OO}{\mathcal{O}}
\newcommand{\rev}{\textit{rev}}
\newcommand{\FIG}[1]{Fig.~\ref{fig:#1}}
\newcommand{\FIGS}[2]{Figs.~\ref{fig:#1} and~\ref{fig:#2}}
\newcommand{\env}{\textit{env}}
\newcommand{\shankaregion}{C2-C5}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\renewcommand{\thesubfigure}{\Alph{subfigure}}
\newcommand{\Author}{Fabio~Zanini and Richard~A.~Neher}
\newcommand{\Title}{Deleterious synonymous mutations hitchhike to high frequency in HIV \env~evolution}
\newcommand{\Keywords}{{HIV}, {synonymous}, {population genetics}}
\hypersetup{pdfauthor={\Author}, pdftitle={\Title}, pdfkeywords={\Keywords}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
\title{\Title}
\author{\Author}
\date{\today}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

HIV evolves rapidly within a single host during the course of the infection.
This evolution is driven by strong selection imposed by the host immune system
via killer T cells (CTLs) and neutralizing antibodies
(ABs)~\citep{rambaut_causes_2004} and facilitated by the high
mutation rate of HIV~\citep{mansky_lower_1995}. When the host develops a CTL or
AB response against a particular HIV epitope, mutations in the viral genome that
reduce or prevent recognition of the epitope frequently emerge. Escape mutations
in epitopes targeted by CTLs typically evolve during early infection and spread
rapidly through the population~\citep{mcmichael_immune_2009}. During chronic
infection, the most rapidly evolving part of the HIV genome are the so called
variable loops of the envelope protein gp120, which need to avoid recognition by
neutralizing ABs.  Mutations in \env, the gene encoding for gp120, spread
through the population within a few months (see \figurename~\ref{fig:aft}, solid
lines). Consistent with this time scale, it is found that serum from a
particular time typically neutralizes virus extracted more than 3-6 month
earlier \citep{richman_rapid_2003}.

These escape mutations are strongly selected for their effect on the amino acid
sequence of the viral proteins. Conversely, synonymous mutations are commonly
used as approximately neutral markers in studies of viral evolution. Neutral
markers are very useful in practice, because there evolution can be compared to
that of putatively functional sites to detect purifying or directional selection
\citep{Bhatt:2011p43255,Hurst:2002p32608,Chen:2004p22606}. In most species,
however, it is  known that synonymous codons don't evolve completely neutrally
but that some codons are favored over others \citep{plotkin_synonymous_2011}.
In addition to maintaining protein function and avoiding immune
recognition, the HIV genome has to ensure efficient processing and translation,
nuclear export, and packaging into the viral capsid: all these processes operate at the RNA
level and are sensitive to synonymous changes. A few functionally important RNA
elements are well characterized. For example, a certain RNA sequence in the HIV
genome, called \rev{} response element (RRE), enhances nuclear export of viral
transcripts~\citep{fernandes_hiv-1_2012}. Another well studied case is the
interaction between viral reverse transcriptase, viral ssRNA, and the host
tRNA$^\text{Lys3}$: the latter is required for priming reverse transcription
(RT) and bound by a specifical pseudoknotted RNA structure in the viral 5'
untranslated region~\citep{barat_interaction_1991, paillart_vitro_2002}.
Nucleotide-level fitness effects have been observed beyond RNA structure as
well. Recent studies have shown that genetically engineered HIV strains with
a skewed codon usage bias towards more or less abundant tRNAs
replicate better or worse \comment{replicate better or produce more protein?},
respectively~\citep{ngumbela_quantitative_2008, li_codon-usage-based_2012}. A similar conclusion has been reached about
influenza, and codon-pessimized influenza strains have been shown to be good
live attenuated vaccines in mice~\citep{mueller_live_2010}. Purifying selection
beyond the protein sequence is therefore expected
\citep{forsdyke_reciprocal_1995,snoeck_mapping_2011}, but it seems reasonable
that the bulk of positive selection through the immune system be restricted to amino acid sequences.

@@

In this paper, we characterize the dynamics of synonymous mutations in \env{}
and show that a substantial fraction of these mutations is deleterious. We
further show that, although such synonymous mutations cannot be used as neutral
markers, the degree to which they hitchhike with nearby nonsynonymous mutations
is very informative both about their on deleterious effect, as well as the
strength and nature of selection for nearby positively selected sites.
Their ability to hitchhike for extended times, which is a core requirement for
our analysis, is rooted in the small recombination rate of
HIV~\citep{neher_recombination_2010, batorsky_estimate_2011}. Extending the
analysis of fixation probabilities to the nonsynonymous mutations, we show that
time dependent selection or strong competition of escape mutations inside the
same epitope are necessary to explain the observed patterns of fixation and
loss.

@@

The central quantity we investigate is the probability of fixation of a
mutation, conditional on its population frequency.  A neutral mutation
segregating at frequency $\nu$ has a probability $P_\text{fix}(\nu) = \nu$ to
spread through the population and fix; in the rest of the cases, i.e.~with
probability $1-\nu$, it goes extinct. As illustrated in the inset of \FIG{aftsyn},
this is a simple consequence of the fact that
(a) exactly one of the $N$ individuals in the current population will be
the common ancestor of the entire future population at a particular locus and
(b) this ancestor has a probability $\nu$ of carrying the mutation (assuming
the neutral mutation is not preferentially associated with genomes of high or
low fitness).
Deleterious or beneficial mutations fix less or
more often than neutral ones, respectively. Time series sequence data enable a
direct observation of both the current frequency $\nu$ of any particular
mutation and its future fate (fixation or extinction).

\subsection{Synonymous polymorphisms in \env, C2-V5, are mostly deleterious}

\FIG{aft} shows time series data of the frequencies of all mutations observed
\env~, C2-V5, in patient p8~\citep{shankarappa_consistent_1999}. Despite many
synonymous mutations reaching high frequency, very few fix
(panel~\ref{fig:aftsyn}); however, many nonsynonymous mutations fix
(panel~\ref{fig:aftnonsyn}).
These observations are quantified in \FIG{fixp}, which stratifies the data
of 7 (resp.~10) patients according to the frequency at which
different mutations are observed (see methods). Considering all mutations in a
frequency interval around $\nu_0$ at some time $t_i$, we calculate the fraction
that is found at frequency 1, at frequency 0, or at intermediate frequency at
later times $t_f$. Plotting these fixed, lost, and polymorphic fraction against
the time interval $t_f-t_i$, we see that most synonymous mutations segregate for
roughly one year and are lost much more frequently than expected (panel
\ref{fig:fixp1}). The long-time probability of fixation versus extinction of
synonymous mutations is shown as a function of the initial frequency $\nu_0$
in panel~\ref{fig:fixp2} (red line). The nonsynonymous seem to follow more or
less the neutral expectation (blue line) -- a point to which we will come back
below.

@@

\citet{bunnik_autologous_2008} present a longitudinal dataset on the entire
\env~gene of 3 patients at $\sim 5$ time points with approximately 5-20
sequences each (see methods). Repeating the above analysis separately on the
C2-V5 region studied above and the remainder of \env~ reveal striking
differences (see \FIG{fixp2}). Within C2-V5, this data fully confirms the
observations made in the data set by \citet{shankarappa_consistent_1999} (red
line). In the remainder of \env, however, observed synonymous mutations behave
neutrally (green line).

These observations suggest that many of the synonymous polymorphisms at
intermediate frequencies in the part of \env~that includes the hypervariable
regions are deleterious, while outside this regions polymorphisms are mostly
roughly neutral. Note that this does not imply that all synonymous mutations are
neutral -- only those mutations observed at high frequencies tend to be neutral.

@@

In addition to RNA secondary structure, we have considered other possible
explanations for a fitness defect of synonymous mutations, in particular codon
usage bias (CUB). HIV is known to prefer A-rich codons over highly expressed
human codons~\citep{jenkins_extent_2003}. Moreover, codon-optimized
and -pessimized viruses have recently been generated and shown to replicate
better or worse than wild type strains,
respectively~\citep{li_codon-usage-based_2012, ngumbela_quantitative_2008,
coleman_virus_2008}. We do not find, however, any evidence for a contribution of
CUB to the ultimate fate of synonymous alleles. Several lines of thought support
this result. First of all, within a single patient, we do not observe any bias
towards more human-like CUB in the synonymous mutations that reach fixation
rather than extinction. Second, although codon-optimized HIV seems to perform
better {\it in vitro}, the distance in CUB between HIV and human genes is not
shrinking at the macroevolutionary level. Third, it is a common phenomenon for
retroviruses to use variously different codons from their hosts, and CUB effects
on fitness are thought to be so small that divergent nucleotide composition has
been suggested as a possible mechanism for viral
speciation~\citep{bronson_nucleotide_1994}. Fourth, CUB in the V1-C5 region is
not very different from other parts of the HIV genome, whereas the reduced
fixation probability is only observed there. In conclusion, although we cannot
exclude an effect of CUB on fitness as a general rule, we expect it to be a
minor effect in our context. \comment{Need to read literature}

@@

While the observation that some fraction of synonymous mutations is deleterious
is not unexpected, it seems odd that we observe them at high population
frequency -- at least in some regions of the genome. The region of \env~ in
which we observe deleterious mutations at high frequency, however, is special in
that it undergoes frequent adaptive changes to evade recognition by neutralizing
antibodies~\cite{williamson_adaptation_2003,richman_rapid_2003}. Due to the limited amount of
recombination in HIV~\cite{neher_recombination_2010,batorsky_estimate_2011},
deleterious mutations that are linked to adaptive variants can reach high
frequency~\citep[genetic hitchhiking,][]{smith_hitch-hiking_1974}.

The potential for hitchhiking is already apparent from the allele frequency
trajectories in \FIG{aft}, where many mutations appear to change rapidly in
frequency as a flock. Deleterious synonymous mutations can be brought to high
frequency by selection on linked nonsynonymous sites, a process
known as {\it genetic draft}~\citep{gillespie_genetic_2000, neher_genetic_2011}. In order
to be advected to high frequency by a linked adaptive mutation, the deleterious
effect of the mutation has to be substantially smaller than the adaptive effect.
The latter was estimated to be on the order of $s_a = 0.01$ per day~\citep{neher_recombination_2010}.
The approximate magnitude of the deleterious effects can be estimated from
\FIG{fixp1}, that shows the distribution of times for synonymous
alleles to reach the fix or get lost starting from intermediate frequencies. The
typical time to loss is of the order of 500 days. If this loss is driven by the
deleterious effect of the mutation, this corresponds to deleterious effects of
roughly $s_d \sim - 0.002$ per day.

@@

The first result of the simulations is that genetic draft can indeed bring weakly
deleterious mutations to high frequencies and results in a dependence of the
fixation probability on initial frequency that is compatible with observations
(see \FIG{simfixpvar}).
To assess the reduction in fixation probability quantitatively, we calculate
the area between the measured fixation probability and the diagonal, which is
the neutral expectation (\FIG{simfixpvar}, lower inset). This area spans a range
between $-0.5$ (no fixation at all), zero (neutral-like curve), and $+0.5$ (always
fixed). Furthermore, we quantify synonymous diversity via the probability
$P_\text{interm}$ of a synonymous derived allele to be at intermediate frequencies
$0.25 < \nu < 0.75$ (\FIG{simfixpvar}, top inset).
The HIV data correspond to an area $A_\text{syn} \sim -0.2$  for synonymous
changes and $A_\text{nonsyn} \sim 0$ for nonsynonymous changes,
and a synonymous diversity $P_\text{interm} \sim 0.005$.
In the three simulations shown, the fixation probability of synonymous alleles
decreases from the neutral expectation ($A_\text{syn} \sim 0$) to zero
($A_\text{syn} \sim -0.5$) as their fitness effect
worsens; the synonymous diversity plummets as well, as it is harder for
deleterious mutations to reach the frequency threshold of 0.25
(\FIG{simfixpvar}, from $-0.0003$ in blue to $-0.003$ in red).

Among all simulations, we select the ones that show HIV-like $A_\text{syn}$ and
$P_\text{interm}$, i.e. a large depression in
fixation probability of synonymous mutations but, simulteneously, a moderately
high synonymous diversity: simulations are gated such that $A_\text{syn} < -0.15$
and $0.0025 < P_\text{interm} < 0.010$, hence allowing an uncertainty on mutation
rates or generation times of about twofold. This selection imposes a strong
constraint on the parameters, in particular on the ones related to synonymous
changes, i.e. their deleterious fraction and effect size (see \FIG{simsfig}).
A high fraction ($\gtrsim 0.8$) of only slightly deleterious synonymous sites,
with effect size $|s_d| \sim 0.002$, is required. This result fits well
the expectation based on the fixation/extinction times above (see \FIG{fixp1}).
Moreover, it corresponds to {\it a priori} theoretical expectations:
(a) since high frequency bins are enriched in neutral mutations (it is easier for
them to hitchhike), the hallmark of deleterious mutations will be visible there
only if they are pervasive; (b) genetic draft stops working if the
deleterious effect size becomes comparable to the adaptive effect size of escape
mutations, because the double mutant has little or no fitness advantage; (c)
mutations with a deleterious effect weaker than approximately $0.001$ are
indistinguishable from neutral ones and are drafted seamlessly to any frequency.

As long as only features of synonymous mutations are filtered, as performed
above, no strong correlation of the two nonsynonymous parameters, effect size
and rate of appearance of epitopes, is observed; in \FIG{simsfig}, the points
shown possess a variety of shapes and colors. In most of these simulations, 
however, nonsynonymous mutations almost always fix once they reach high frequencies
-- their nonsynonymous fixation area is well above zero. This is incompatible
with the blue line in \FIG{fixp}: in an HIV infection, nonsynonymous mutations fix as if they
were neutral. Although there are no doubts that nonsynonymous variation in the
variable regions is driven by positive selection and not by genetic drift, the
biological reason for the neutral-like curve in \FIG{fixp} is unclear.
Inspecting the trajectories of nonsynonymous mutations
suggests the rapid rise and fall of many alleles. We test two possible
mechanisms that are biologically plausible and could explain the transient rise
of nonsynonymous mutations: time-dependent selection and within-epitope
competition.

The former explanation can be formulated as follows: if the immune system
recognizes the escape mutant before its fixation, the mutant might cease to be
beneficial and disappear soon, despite its quick initial rise in frequency. In support of this idea,
\citet{richman_rapid_2003, bunnik_autologous_2008} report antibody responses to
escape mutants. These respones are delayed by a few months, roughly matching the
average time needed by an escape mutant to cross frequencies of order one.
In the alternative explanation, several different escape
mutations within the same epitope might arise almost simultaneously and start to
spread. Their benefits are not additive, because each of them is
essentially sufficient to escape. As a consequence, several escape mutations rise to
high frequency rapidly, while the one with the smallest cost in terms of replication,
packaging, etc.~is most likely to eventually fix. The emergence of
multiple sweeping nonsynonymous mutations in real HIV infections has been shown
previously~\citep{moore_limited_2009, bar_early_2012}.
We tested both hypotheses in our simulation framework and they both seem to
achieve a reduced, neutral-like fixation probability while maintaining a high
rate of substitutions, compatibly with HIV measurements. In real HIV infections,
both mechanisms are likely to be playing a role.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ARTICLE ABOUT FATE OF SYNONYMOUS MUTATIONS IN HIV
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[rmp, twocolumn]{revtex4}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage{amsmath,amsfonts,amssymb,eucal,eurosym,textcomp}
\usepackage{color}
\usepackage{graphicx}
\usepackage[caption=false]{subfig}
\usepackage{natbib}
\usepackage{pslatex}
\usepackage[colorlinks,linkcolor=red,citecolor=red]{hyperref}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\graphicspath{{./figures/}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\comment}[1]{\textit{\textcolor{red}{#1}}}
\newcommand{\mut}{\mu}
\newcommand{\mfit}{\langle F\rangle}
\newcommand{\mexpfit}{\langle e^{F}\rangle}
\newcommand{\pfix}{P_{fix}}
\newcommand{\ox}{r}
\newcommand{\co}{\rho}
\newcommand{\gt}{g}
\newcommand{\locus}{s}
\newcommand{\locuspm}{t}
\newcommand{\OO}{\mathcal{O}}
\newcommand{\rev}{\textit{rev}}
\newcommand{\FIG}[1]{Fig.~\ref{fig:#1}}
\newcommand{\FIGS}[2]{Figs.~\ref{fig:#1} and~\ref{fig:#2}}
\newcommand{\env}{\textit{env}}
\newcommand{\shankaregion}{C2-C5}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\renewcommand{\thesubfigure}{\Alph{subfigure}}
\newcommand{\Author}{Fabio~Zanini and Richard~A.~Neher}
\newcommand{\Title}{Deleterious synonymous mutations hitchhike to high frequency in HIV \env~evolution}
\newcommand{\Keywords}{{HIV}, {synonymous}, {population genetics}}
\hypersetup{pdfauthor={\Author}, pdftitle={\Title}, pdfkeywords={\Keywords}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
\title{\Title}
\author{\Author}
\date{\today}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

HIV evolves rapidly within a single host during the course of the infection.
This evolution is driven by strong selection imposed by the host immune system
via killer T cells (CTLs) and neutralizing antibodies
(ABs)~\citep{rambaut_causes_2004} and facilitated by the high
mutation rate of HIV~\citep{mansky_lower_1995,abram_nature_2010}. When the host
develops a CTL or AB response against a particular HIV epitope, mutations in the viral genome that
reduce or prevent recognition of the epitope frequently emerge. Escape mutations
in epitopes targeted by CTLs typically evolve during early infection and spread
rapidly through the population~\citep{mcmichael_immune_2009}. During chronic
infection, the most rapidly evolving part of the HIV genome are the so called
variable loops of the envelope protein gp120, which need to avoid recognition by
neutralizing ABs.  Mutations in \env, the gene encoding for gp120, spread
through the population within a few months (see \figurename~\ref{fig:aft}, solid
lines). Consistent with this time scale, it is found that serum from a
particular time typically neutralizes virus extracted more than 3-6 month
earlier \citep{richman_rapid_2003}.

These escape mutations are strongly selected for their effect on the amino acid
sequence of the viral proteins. Conversely, synonymous mutations are commonly
used as approximately neutral markers in studies of viral evolution. Neutral
markers are very useful since their dynamics can be compared to
that of putatively functional sites to detect purifying or directional selection
\citep{Bhatt:2011p43255,Hurst:2002p32608,Chen:2004p22606}. In most species,
however, it is  known that synonymous codons don't evolve completely neutrally
but that some codons are favored over others \citep{plotkin_synonymous_2011}.
This is expected to be particularly important in compact viruses like HIV. 
In addition to maintaining protein function and avoiding the adaptive immune
recognition, the HIV genome has to ensure efficient processing and translation,
nuclear export, and packaging into the viral capsid: all these processes operate at the RNA
level and are sensitive to synonymous changes. A few functionally important RNA
elements are well characterized. For example, a certain RNA sequence in the HIV
genome, called \rev{} response element (RRE), enhances nuclear export of
full length or partially spliced viral transcripts~\citep{fernandes_hiv-1_2012}.
Another well studied case is the interaction between viral reverse transcriptase, viral ssRNA, and the host
tRNA$^\text{Lys3}$: the latter is required for priming reverse transcription
(RT) and is bound by a pseudoknotted RNA structure in the viral 5'
untranslated region~\citep{barat_interaction_1991, paillart_vitro_2002}.
Nucleotide-level fitness effects have been observed beyond RNA structure as
well. Recent studies have shown that genetically engineered HIV strains with
altered codon usage can in some cases produce more viral protein, but in general
replicate less efficiently~\citep{ngumbela_quantitative_2008,
li_codon-usage-based_2012,keating_rich_2009}.
Codon-deoptimization has been suggested as attenuation strategy for polio and 
influenza~\citep{mueller_live_2010,coleman_virus_2008}. Purifying
selection beyond the protein sequence is therefore expected
\citep{forsdyke_reciprocal_1995,snoeck_mapping_2011}, but while positive
selection through the immune system be mainly restricted to changes in the
amino acid sequence.

@@

In this paper, we characterize the dynamics of synonymous mutations in \env{}
and show that a substantial fraction of these mutations is deleterious. We then
compare our observations to computational models of HIV evolution and derive
estimates for the effect synonymous mutations have on fitness.
Hitchhiking of synonymous mutations with escape mutations is due to the small
recombination rate of HIV~\citep{neher_recombination_2010,
batorsky_estimate_2011}. Extending the analysis of fixation probabilities to the
nonsynonymous mutations, we show that time dependent selection or strong
competition of escape mutations inside the same epitope are necessary to explain
the observed patterns of fixation and loss.

@@

The central quantity we investigate is the probability of fixation of a
mutation, conditional on its population frequency.  A neutral mutation
segregating at frequency $\nu$ has a probability $P_\text{fix}(\nu) = \nu$ to
spread through the population and fix; in the rest of the cases, i.e.~with
probability $1-\nu$, it goes extinct. As illustrated in the inset of \FIG{aftsyn},
this is a simple consequence of the fact that
(a) exactly one of the $N$ individuals in the current population will be
the common ancestor of the entire future population at a particular locus and
(b) this ancestor has a probability $\nu$ of carrying the mutation (assuming
the neutral mutation is not preferentially associated with genomes of high or
low fitness).
Deleterious or beneficial mutations fix less or
more often than neutral ones, respectively. \FIG{aft}A\&B show 
the time course the frequencies of all synonymous and non-synonymous mutations
observed \env~, C2-V5, in patient p8~\citep{shankarappa_consistent_1999},
respectively. Despite many synonymous mutations reaching high frequency, very
few fix (panel~\ref{fig:aftsyn}); however, many nonsynonymous mutations fix
(panel~\ref{fig:aftnonsyn}).

\subsection{Synonymous polymorphisms in \env, C2-V5, are mostly deleterious}
We study the dynamics and fate of synonymous mutations more quantitatively by
analyzing data from 7 patients from
\citet{shankarappa_consistent_1999,liu_selection_2006} and 3 patients from
\citet{bunnik_autologous_2008} (see methods). The data set from
\citet{shankarappa_consistent_1999,liu_selection_2006} is restricted to the
C2-V5 region of \env, while \citet{bunnik_autologous_2008} covers the
majority of \env. Considering all mutations in a
frequency interval around $\nu_0$ at some time $t_i$, we calculate the fraction
that is found at frequency 1, at frequency 0, or at intermediate frequency at
later times $t_f$. Plotting these fixed, lost, and polymorphic fraction against
the time interval $t_f-t_i$, we see that most synonymous mutations segregate for
roughly one year and are lost much more frequently than expected (panel
\ref{fig:fixp1}). The long-time probability of fixation versus extinction of
synonymous mutations is shown as a function of the initial frequency $\nu_0$ in
panel~\ref{fig:fixp2} (red line). Restricted to the region C2-V5, we find that
$\pfix$ of synonymous mutations is far below the neutral expectation.
Outside of C2-V5 using data from \citet{bunnik_autologous_2008} only, no such
reduction in $\pfix$ is found. Restricted to the C2-V5 area, the data from
\citet{bunnik_autologous_2008} is fully compatible with data from
\citet{shankarappa_consistent_1999}. The nonsynonymous seem to follow more or
less the neutral expectation (blue line) -- a point to which we will come back below.

@@

When interpreting these results for the fixation probabilities, it is important
to distinguish random mutations, and polymorphisms at a given frequency that has
already been filtered by selection.
A polymorphism could be beneficial to the virus and on its way to fixation. In
this case, we expect that it fixes almost surely given we see it at high
frequency. If, on the other hand, the polymorphism is deleterious it must have
gotten to high frequency by a chance event (genetic drift or hitch-hiking), and
we expect that selection will drive it out of the population again. Hence our
observations suggest that many of the synonymous polymorphisms at intermediate
frequencies in the part of \env~that includes the hypervariable regions are
deleterious, while outside this regions polymorphisms are mostly roughly
neutral. Note that this does not imply that all synonymous mutations are neutral
-- only those mutations observed at high frequencies tend to be neutral.

@@

In addition to RNA secondary structure, we have considered other possible
explanations for a fitness defect of synonymous mutations, in particular codon
usage bias (CUB). HIV is known to prefer A-rich codons over highly expressed
human codons~\citep{jenkins_extent_2003,kuyl_biased_2012}. We do not find,
however, any evidence for a contribution of CUB to the ultimate fate of
synonymous alleles consistent with the fact that CUB between HIV and human genes
is not shrinking at the macroevolutionary level \citep{kuyl_biased_2012}. 
Furthermore, CUB in the V1-C5 region is not very different from other parts of
the HIV genome, whereas the reduced fixation probability is only observed there. In
conclusion, although we cannot exclude an effect of CUB on fitness as a general
rule, we expect it to be a minor effect in our context.

@@

While the observation that some fraction of synonymous mutations is deleterious
is not unexpected, it seems odd that we observe them at high population
frequency -- at least in some regions of the genome. The region of \env~ in
which we observe deleterious mutations at high frequency, however, is special in
that it undergoes frequent adaptive changes to evade recognition by neutralizing
antibodies~\cite{williamson_adaptation_2003,richman_rapid_2003}. Due to the limited amount of
recombination in HIV~\cite{neher_recombination_2010,batorsky_estimate_2011},
deleterious mutations that are linked to adaptive variants can reach high
frequency. This process is known as hitch-hiking \citep{smith_hitch-hiking_1974}
or genetic draft \citep{gillespie_genetic_2000,neher_genetic_2011}.

The potential for hitchhiking is already apparent from the allele frequency
trajectories in \FIG{aft}, where many mutations appear to change rapidly in
frequency as a flock. In order
to be advected to high frequency by a linked adaptive mutation, the deleterious
effect of the mutation has to be substantially smaller than the adaptive effect.
The latter was estimated to be on the order of $s_a = 0.01$ per day~\citep{neher_recombination_2010}.
The approximate magnitude of the deleterious effects can be estimated from
\FIG{fixp1}, that shows the distribution of times for synonymous
alleles to reach the fix or get lost starting from intermediate frequencies. The
typical time to loss is of the order of 500 days. If this loss is driven by the
deleterious effect of the mutation, this corresponds to deleterious effects of
roughly $s_d \sim - 0.002$ per day.

@@

The first result of the simulations is that genetic draft can indeed bring weakly
deleterious mutations to high frequencies and results in a dependence of the
fixation probability on initial frequency that is compatible with observations
(see \FIG{simfixpvar}).
To assess the reduction in fixation probability quantitatively, we calculate
the area between the measured fixation probability and the diagonal, which is
the neutral expectation (\FIG{simfixpvar}, lower inset). This area spans a range
between $-0.5$ (no fixation at all), zero (neutral-like curve), and $+0.5$ (always
fixed). Furthermore, we quantify synonymous diversity via the probability
$P_\text{interm}$ of a synonymous derived allele to be at intermediate frequencies
$0.25 < \nu < 0.75$ (\FIG{simfixpvar}, top inset).
The HIV data correspond to an area $A_\text{syn} \sim -0.2$  for synonymous
changes and $A_\text{nonsyn} \sim 0$ for nonsynonymous changes,
and a synonymous diversity $P_\text{interm} \sim 0.005$.
In the three simulations shown in \FIG{simfixpvar}, the fixation probability of
synonymous alleles decreases from the neutral expectation ($A_\text{syn} \sim 0$) to zero
($A_\text{syn} \sim -0.5$) as their fitness effect
worsens; the synonymous diversity plummets as well, as it is harder for
deleterious mutations to reach the frequency threshold of 0.25.

To map the parameter range of the model that is compatible with the data, we
repeatedly simulated the evolution with random choices for the deleterious
effect of synonymous mutations, the fraction of synonymous mutations that are
deleterious, the escape rate of adaptive mutations, and the frequency of
escapes.  Among all simulations, we select the ones that show $A_\text{syn}$ and
$P_\text{interm}$ as observed in the data, i.e., a large depression in fixation
probability of synonymous mutations but, simulteneously, a moderately high
synonymous diversity. Specifically, \FIG{simsfig} shows parameter combinations
for which we found $A_\text{syn} < -0.15$ and $0.0025 < P_\text{interm} < 0.010$
(twofold variation around the observed value). This selection imposes a strong
constraint on the deleterious fraction and effect size (see \FIG{simsfig}).
A high fraction ($\gtrsim 0.8$) of only slightly deleterious synonymous sites,
with effect size $|s_d| \sim 0.002$, is required. This result fits well the
expectation based on the fixation/extinction times above (see \FIG{fixp1}).
The results are plausible:
(a) since high frequency bins are enriched in neutral mutations (it is easier
for them to hitchhike), a depression in $\pfix$ will be
visible only if deleterious mutations are pervasive; (b) in order to
hitchhike, the deleterious effect size has to be less than the escape rate, otherwise the
double mutant has little or no fitness advantage; (c) mutations with a 
deleterious effect weaker than approximately $0.001$ behave neutrally
consistent with the typical coalescent times observed in HIV.

As long as only features of synonymous mutations are filtered, as performed
above, no strong correlation of the two nonsynonymous parameters, escape rate
and rate of appearance of epitopes, is observed; in \FIG{simsfig}, the points
shown possess a variety of shapes and colors. In most of these simulations, 
however, nonsynonymous mutations almost always fix once they reach high frequencies
-- their nonsynonymous fixation area is well above zero. This is incompatible
with the blue line in \FIG{fixp}: in an HIV infection, nonsynonymous mutations fix as if they
were neutral. Although there are no doubts that nonsynonymous variation in the
variable regions is driven by positive selection and not by genetic drift, the
biological reason for the neutral-like curve in \FIG{fixp} is unclear.
Inspecting the trajectories of nonsynonymous mutations
suggests the rapid rise and fall of many alleles. We test two possible
mechanisms that are biologically plausible and could explain the transient rise
of nonsynonymous mutations: time-dependent selection and within-epitope
competition.

The former hypothesis can be formulated as follows: if the immune system
recognizes the escape mutant before its fixation, the mutant might cease to be
beneficial and disappear soon, despite its quick initial rise in frequency. In support of this idea,
\citet{richman_rapid_2003, bunnik_autologous_2008} report antibody responses to
escape mutants. These respones are delayed by a few months, roughly matching the
average time needed by an escape mutant to cross frequencies of order one.
In the alternative hypothesis, several different escape
mutations within the same epitope might arise almost simultaneously and start to
spread. Their benefits are not additive, because each of them is
essentially sufficient to escape. As a consequence, several escape mutations rise to
high frequency rapidly, while the one with the smallest cost in terms of replication,
packaging, etc.~is most likely to eventually fix. The emergence of
multiple sweeping nonsynonymous mutations in real HIV infections has been shown
previously~\citep{moore_limited_2009, bar_early_2012}.
We tested both hypotheses in our simulation framework and they both seem to
achieve a reduced, neutral-like fixation probability while maintaining a high
rate of substitutions, compatibly with HIV measurements \comment{Supplementary}.
In real HIV infections, both mechanisms are likely to be playing a role.
